---
title: "Blake Davenport Case Study"
author: "Blake Davenport"
date: "June 20, 2016"
output: html_document
---

#Introduction
###The following work will examine two data sets. The first set is GDP data for 190 ranked countries. 
###The second data set is educational data for most of the same 190 countries. The software package
###R version 3.3.0 (2016-05-03) -- "Supposedly Educational"Copyright (C) 2016 
###The R Foundation for Statistical Computing Platform:x86_64-w64-mingw32/x64 (64-bit) will be used for the data analysis. 
###This is my first formal attempt at using R software, and I found the experience interesting.

##I Used The Following RLibraries For This Case Study:
###library(data.table)
###library(dplyr)
###library(ggplot2)
```{r}
library(data.table)
library(dplyr)
library(ggplot2)
```
##The first step was to download the GDP and Educational Data
##I did that from the links provided and then ran the files 
##from my machine
```{r}
GDP<-read.csv("GDP.csv")
ED<-read.csv("EDSTATS_Country.csv")
list.files()
```
###This series of code identified the URL for the data sets, I downloaded the files to my computer, and then confirmed the files were on my machine.
###These commands have characters read as strings instead of factors. 
###As I learned throughout the effort, it is important when working with R to have the data read properly by the software.

###Review of GDP Data
```{r}
GDPv1<-read.csv("GDP.csv", stringsAsFactors=FALSE, header=FALSE)
str(GDPv1)
head(GDPv1)
tail(GDPv1)
```
###The following code checks for NAs in the GDPv1 data
```{r}
sum(!is.na(GDPv1[,c(1:5)]))
```
###This series of code provides a base description of the GDP data. 
###It showed the GDP data 243 observations of six variables. Variable 3 returned all NA. First review 
###indicates only V2, V4, and V5 contain useful data.
###Empty space composed the first five rows of the GDP data, and all useful information concludes by 
###row 199. 
```{r}
GDPv2<-GDPv1[c(6:200), 1:5]#I got this idea from: http://www.ats.ucla.edu/stat/r/faq/subset_R.htm 
GDPv3<-GDPv2[, c(1, 2, 4, 5)]
names(GDPv3)<-c("CountryCode", "GDPRanking", "Nation", "GDPInDollars")#This will help merging the two data sets
head(GDPv3)
```
###To confirm the columns are labeled correctly
```{r}
GDPv3$GDPInDollars<-gsub(",","",GDPv3$GDPInDollars, fixed = TRUE)
GDPv3$GDPInDollars<-gsub(".","",GDPv3$GDPInDollars, fixed = TRUE)
```
###This removes the commas and periods in the GDPInDollars column
```{r}
GDPv3$GDPInDollars<-as.numeric(GDPv3$GDPInDollars)
GDPv3$GDPInDollars<-as.integer(GDPv3$GDPInDollars)
```
###These two lines of code will reappear. I think there were times I may have created a matrix instead
###or a data table. Using this code ensured the columns were read as numeric and integers
```{r}
GDPv4<-subset(GDPv3, GDPv3$CountryCode!="")#This removes the data gaps from row 200 to 233

GDPFinal<-GDPv4[,c(1,2,4)]#This pulls only the data required to answer the case study questions
str(GDPFinal)
head(GDPFinal)
tail(GDPFinal)
GDPFinal<-GDPFinal[order(GDPFinal$CountryCode),]
```
# This code orders the data by CountryCode instead of by GDP

###In general, the same process applied to the GDP data will now be done to the EDSTATS data.
```{r}
EDv1<-read.csv("EDSTATS_Country.csv", stringsAsFactors=FALSE, header=FALSE)
str(EDv1)
head(EDv1)
tail(EDv1)
sum(!is.na(EDv1[,c(1, 2, 9)]))#The total number of NAs
```
###This series of code provides a base description of the EDSTATS data. 
###EDSTATS data has 242 observations of 32 variables. V1, V2, V4, and V9 appear to contain the most 
###useful data. Many of the variables do not appear relevant to the case study. Variable 32 returned all 
###NAs. Useful data concludes on row 242 with Zimbabwe
```{r} 
EDv2<-EDv1[c(2:242), 1:9]# Same step as done to the GDP data
EDFinal<-EDv2[,c(1, 2, 9)]
names(EDFinal)<-c("CountryCode", "Nation", "IncomeGroup")
head(EDFinal)
```
## Merging Income and GDP data
```{r}
DataGroupv1 <- merge(EDFinal, GDPFinal, by = "CountryCode", all = TRUE)
str(DataGroupv1) # Review structure details
head(DataGroupv1) # Looking for blank observations or NA
DataGroupv2<-subset(DataGroupv1, DataGroupv1$GDPRanking!="")# To remove blank spaces
sum(!is.na(DataGroupv2))#shows there are still 975 instances of NA
```
#Question 1: Match the data based on the country shortcode. How many of the IDs match?
##How many IDs match?
```{r}
length(intersect(GDPFinal$CountryCode, EDFinal$CountryCode))
```
##Remove rows with missing data
```{r}
DataGroupv3 <- DataGroupv2[complete.cases(DataGroupv2),]
nrow(DataGroupv3) # Provide row count after removing rows with missing data
```
#Question 2: Sort the data frame in ascending order by GDP (so United States is last). 
What is the 13th country in the resulting data frame?
```{r}
DataGroupv2 <- DataGroupv2[order(DataGroupv2$GDPInDollars, decreasing = FALSE),] # This sorts the data frame by GDP
DataGroupv2$Nation[13] # Display only the 13th country in the data frame
```
#Question 3: What are the average GDP rankings for the "High income: OECD" and "High income: nonOECD" groups?
```{r}
DataGroupv2 <- DataGroupv2[order(DataGroupv2$IncomeGroup, decreasing = FALSE),]
```
###for nonOECD
```{r}
nonOECD<- DataGroupv2[c(1:29), 4]
nonOECD<-as.numeric(nonOECD)
nonOECD<-as.integer(nonOECD)
mean(nonOECD, na.rm=TRUE)
```
###For OECD
```{r}
OECD<-DataGroupv2[c(30:60), 4]
OECD<-as.numeric(nonOECD)
OECD<-as.integer(nonOECD)
mean(OECD, na.rm=TRUE)
```
###I knew and found various ways to get the mean as requested. However, I continued to receive error
###messages each time. The code above was a rather brute force method to obtain an answer. It also 
###illustrates the numeric, integer, and matrix/data table issues briefly discussed above.

#Question 4: Plot the GDP for all of the countries. Use ggplot2 to color your plot by IncomeGroup:http://ucsdlib.github.io/win2016-gps-intro-R/ggplot.html
```{r}
ggplot(data=DataGroupv2, aes(x=IncomeGroup, y=GDPRanking, color=IncomeGroup)) + geom_boxplot() + geom_smooth(method="lm", size=.5) 
```
###I tried a logarithmic adjustments and to rescale the y axis labeling multiple times and multiple way. 
###None of those attempts worked. I am still not sure where my error is on these steps. My first 
###attempt also tried to scatter plot the data, but it did not produce good results. The boxplot is only slightly better.

#Question 5: Cut the GDP ranking into 5 separate quantile groups. 
##Make a table versus Income Group. How many countries are lower middle income but among the 38 nations with the highest GDP?
```{r}
DataGroupv3<-DataGroupv2
DataGroupv3$GDPQuantile<-ntile(DataGroupv3$GDPRanking, 5)
table(DataGroupv3$GDPRanking, DataGroupv3$GDPQuantile, dnn=c("IncomeGroup", "DataGroupv3"))
sum(DataGroupv3[(nrow(DataGroupv3)-37):nrow(DataGroupv3),]$IncomeGroup=="Lower middle income")
```
###I'm not sure I fulfilled the intent of the question with this code, but it does answer the question

#Conclusion
###I encountered a number of issues during this project that taught me some valuable lessons.
###I am most comfortable is Excel, and I slowly started using R like Excel when I could not
###find suitable solutions. I do not know if that is the proper thought process at this time
###but as is common when learning any new language, I tend to revert to what I know.